\section{Quantum Logic Gates and Circuits}
Just as classical algorithms on its basis are constructed with logic gates like AND, OR, NOT and so on, 
quantum algorithms are built using \textit{quantum circuit model}.
%
%
In quantum computations we apply a sequence of operations, known as quantum logic gates, to a set of initialized qubits.
%
%
The computation then finishes with a measurement to extract the final state and gives us the desired result \cite{deutsch1989quantum}.
%
%
A main rule of quantum systems is that their evolution in time is deterministic and reversible, 
because of that, each quantum gate must be an invertible operation.
%
%
The only exception to that rule is the measurement operation, which irreversibly collapses the state and provides our result.
%
%
Additionally, because these gates manipulate probability amplitudes, they must preserve the fundamental rule that all probabilities sum exactly to 100\%.
To satisfy these twin requirements of reversibility and probability conservation, all quantum gates are mathematically represented as \textit{unitary matrices} \cite{nielsen2010quantum}.
%
%
\subsection{Single-Qubit Gates}
Gates for a single qubit operate on the state space $\mathbb{C}^2$, and can be represented as $2 \times 2$ unitary matrices.
%
%
The most known and useful of gates for a single quibit are three Pauli matrices.
They correspond to $180^{\circ}$ degree ($\pi$ radian) rotations around each of axes ($x$, $y$, and $z$) of the Bloch sphere:
%
%
\begin{align}
    X &= \begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix} \\
    Y &= \begin{pmatrix} 0 & -i \\ i & 0 \end{pmatrix} \\
    Z &= \begin{pmatrix} 1 & 0 \\ 0 & -1 \end{pmatrix}
\end{align}
%
%
The $X$ gate is the equivalent of the NOT gate in quantum computing. 
It flips the sate of the quibit which maps $\ket{0} \to \ket{1}$ and $\ket{1} \to \ket{0}$.
%
%
The $Z$ gate on the other hand is like a phase flip.
It leaves the $\ket{0}$ state unchanged, but flips the sign of $\ket{1}$, mapping it to $-\ket{1}$.
%
%
Gate $Y$ works on a quibit as gate $X$ fliping the state of the quibit but also applyes the complex phase shift to it mapping 
$\ket{0} \to i\ket{1}$ and $\ket{1} \to -i\ket{0}$.
%
%
Probably the most common operation amongs the operations on a single qubit is the Hadamard gate, denoted $H$.
It is a basic gate for generating quantum superposition from the input state and typically occurs right on the start of the circuit.
It creates a equal probability distribution between $\ket{0}$ and $\ket{1}$ state and can be denoted as:
%
%
\begin{equation}
    H = \frac{1}{\sqrt{2}}\begin{pmatrix} 1 & 1 \\ 1 & -1 \end{pmatrix}
\end{equation}
%
%
Applying $H$ to $\ket{0}$ yields the positive superposition state $\ket{+} = \frac{1}{\sqrt{2}}(\ket{0} + \ket{1})$ and 
applying it to $\ket{1}$ yields the negative superposition state $\ket{-} = \frac{1}{\sqrt{2}}(\ket{0} - \ket{1})$ \cite{mermin2007quantum}.
%
%
Other useful operations are the Phase gate ($S$) and the $\pi/8$ gate ($T$).
Instead of flipping the state (like the $X$ gate), these gates rotate it around the Z-axis (similarly to the $Z$), 
changing the relative phases without changing the measurement probabilities. 
%
%
This gate can be denoted as:
\begin{equation}
    S = \begin{pmatrix} 1 & 0 \\ 0 & i \end{pmatrix}, \quad T = \begin{pmatrix} 1 & 0 \\ 0 & e^{i\pi/4} \end{pmatrix}
\end{equation}
%
%
\subsection{Two Qubit Gates: CNOT, SWAP, and Controlled-Phase}
While gates that operates on single qubit can modyfy it state spaces (probabilities and phase), 
we must now be able to apply gates operating on multiple qubits to be able to perform conditional logic or generate entanglement.
%
%
The basic operation on two qubits is the Controlled-NOT (CNOT) gate.
%
%

We can say, that the CNOT gate works with a control qubit and a target qubit.
If the control qubit is in the $\ket{1}$ state, the gate applies the $X$ gate to the target qubit (it flips a bit of target qubit).
If the control qubit is in the $\ket{0}$ state, the target qubit is left completely untouched.

%
%
In the standard computational basis $\{\ket{00}, \ket{01}, \ket{10}, \ket{11}\}$, 
where the first qubit acts as the control, the matrix representation of CNOT looks like:
%
%
\begin{equation}
    \text{CNOT} = \begin{pmatrix} 
    1 & 0 & 0 & 0 \\ 
    0 & 1 & 0 & 0 \\ 
    0 & 0 & 0 & 1 \\ 
    0 & 0 & 1 & 0 
    \end{pmatrix}
\end{equation}

The CNOT gate is really useful and will serves as the basic mechanism for entangling previously independent qubits.
For example, if we start with the state $\ket{00}$, 
then we apply a Hadamard gate to the first qubit to create a superposition state with equal probabilities, 
and then apply a CNOT gate, 
we generate the entangled Bell state $\ket{\Phi^+} = \frac{1}{\sqrt{2}}(\ket{00} + \ket{11})$ \cite{nielsen2010quantum}.

%
%
Next really practical and usefull operation on two quibits is the SWAP gate. It exchanges the quantum states of two distinct qubits.
%
%

This gate is usefull for routing information across physical hardware topologies, 
as not all qubits are directly connected to each other. 
Typically only the nearest ones are only connected so we have to change theirs positions.
It also will act as the final bit reversal step of the Quantum Fourier Transform (QFT) used in Shor's algorithm:
\begin{equation}
    \text{SWAP} = \begin{pmatrix} 
    1 & 0 & 0 & 0 \\ 
    0 & 0 & 1 & 0 \\ 
    0 & 1 & 0 & 0 \\ 
    0 & 0 & 0 & 1 
    \end{pmatrix}
\end{equation}

While the CNOT conditionally a flip a bit or leaves it unchanged, 
the Controlled-Phase gate, denoted as $CP(\theta)$ or $CR_z(\theta)$ applies a conditional phase shift.
%
%
This gate adds a complex phase of $e^{i\theta}$ to the system if, and only if, 
both the control and target qubits are simultaneously in the $\ket{1}$ state:
%
%
\begin{equation}
    CP(\theta) = \begin{pmatrix} 
    1 & 0 & 0 & 0 \\ 
    0 & 1 & 0 & 0 \\ 
    0 & 0 & 1 & 0 \\ 
    0 & 0 & 0 & e^{i\theta} 
    \end{pmatrix}
\end{equation}

This gate enables the precise extraction of periodic phases, which will necessary for tasks like integer factorization (Shor's algorithm).
%
%

\subsection{Advanced and Multi-Controlled Gates}

To create more complex logical structures, like the Search Oracles in Grover's algorithm or 
the modular arithmetic in Shor's algorithm. We need gates that condition their behavior on several qubits at once.

%
%
The Toffoli gate, also referred to as the Controlled-Controlled-NOT (CCNOT), is the most famous example of this.
It works quite simple. It observes two control qubits and applies an $X$ gate to a single target qubit if, and only if, 
both controls are in the $\ket{1}$ state.
%
%
Because the Toffoli gate is proven to be universal for classical reversible computation, 
it is the key gate for building quantum oracles that will evaluate classical boolean functions \cite{nielsen2010quantum}. 
The CCNOT gate can be denoted as:
%
%
\begin{equation}
    \text{CCNOT} = \begin{pmatrix} 
    1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
    0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
    0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\
    0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
    0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 \\
    0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 
    \end{pmatrix}
\end{equation}

%
%
For more advanced quantum algorithms, the Toffoli gate is often generalized into the Multi-Controlled X gate ($MCX$).
The $MCX$ gate flips a target qubit conditioned on the state of an arbitrary number of $n$ control qubits.

%
%
Similarly, the diffusion step in Grover's algorithm relies heavily on a Multi-Controlled Z gate ($MCZ$), which is 
instead of flipping the bit (applying the $X$ gate), the $MCZ$ gate flips the phase (the sign, by applying the $Z$ gate), 
but only when all $n$ qubits in the same time are in the $\ket{1}$ state.

%
%
In practice we may have not ready native $MCZ$ gate availbale to us, but we can easily synthesize one by taking an $MCX$ 
gate and wrapping its target qubit in Hadamard gates.
This trick smoothly shifts the bit flip operation into a phase flip operation: 
$MCZ = (I \otimes \dots \otimes H) MCX (I \otimes \dots \otimes H)$ \cite{barenco1995elementary}.

%
%
Many algorithms may also rely on the generalized Controlled-$U$ operation ($c-U$).
Imagine an arbitrary quantum operation $U$ designed to act on a large target set of $n$ qubits.
The $c-U$ gate check if a single control qubit is equal to $\ket{1}$, and if so, 
it triggers the $U$ operation on the target qubits.