\section{Mathematical Prerequisites}

To get deeper into quantum algorithms, one must first understand in the mathematical language of quantum mechanics, which is linear algebra.
%
%

Unlike classical computing systems, whose states are described by discrete deterministic variables, quantum systems are probabilistic.
Which causes, that their foundational behavior is formalized using complex vector spaces, inner products, and linear operator theory \cite{nielsen2010quantum}.
%
%
\subsection{Hilbert Spaces and Dirac Notation}

The state of any isolated quantum system is completely described by a unit vector residing in a complex \textit{Hilbert space}, 
denoted by $\mathcal{H}$.
%
%
Hilbert space can be thought of as a generalization of a standard Euclidean space.
The familiar geometric notions of length and angle still apply, but the space is extended to accommodate complex numbers.
%
%
For the finite dimension systems relevant to quantum computing one, such as a register of $n$ interacting qubits, 
the Hilbert space is isomorphic to $\mathbb{C}^{2^n}$.
This means that it is $2^n$ dimensional complex vector space and the exponential growth factor 
hints at the parallel processing power of quantum computing.
%
%

To manipulate these state vectors, we use Dirac's \textit{bra-ket} notation \cite{dirac1939new}.
It's provides a algebraic shortcut that maps perfectly to the underlying linear algebra operations:
%
%
\begin{itemize}
    \item \textbf{Ket} ($\ket{\psi}$): A standard column vector representing a distinct quantum state.
    \item \textbf{Bra} ($\bra{\psi}$): The Hermitian conjugate (conjugate transpose) of the ket, representing a row vector residing in the dual space $\mathcal{H}^*$.
\end{itemize}
%
%

For a vector $\psi \in \mathbb{C}^d$, we write:
\begin{equation}
    \begin{split}
        \ket{\psi} &= \begin{pmatrix} z_1 \\ \vdots \\ z_d \end{pmatrix}, \\
        \bra{\psi} &= \ket{\psi}^\dagger = \begin{pmatrix} z_1^* & \dots & z_d^* \end{pmatrix}
    \end{split}
\end{equation}
where, $z^*$ is the complex conjugate of $z$.
%
%
\subsection{Inner and Outer Products}

The \textbf{inner product} of two state vectors $\ket{\phi}$ and $\ket{\psi}$ is a scalar $c \in \mathbb{C}$, 
denoted by the combined bracket $\braket{\phi}{\psi}$.
%
%
This operation measures the "overlap" or similarity between two quantum states and it satisfies the following properties:
%
%
\begin{enumerate}
    \item Conjugate symmetry: $\braket{\phi}{\psi} = \braket{\psi}{\phi}^*$.
    \item Linearity in the second argument: $\bra{\phi} (a\ket{\psi_1} + b\ket{\psi_2}) = a\braket{\phi}{\psi_1} + b\braket{\phi}{\psi_2}$.
    \item Positive definiteness: $\braket{\psi}{\psi} \geq 0$, with equality if and only if $\ket{\psi} = 0$.
\end{enumerate}
%
%

Two vectors are considered \textit{orthogonal} when their inner product is equal to 0 ($\braket{\phi}{\psi} = 0$). 
%
%
This condition signifies that they represent mutually exclusive quantum states.
%
%
The norm corresponds to the geometric length of a state vector and it is defined as $||\psi|| = \sqrt{\braket{\psi}{\psi}}$.
%
%
For any vector to represent a valid quantum state, 
it must be properly normalized to unity, which mean that it's norm have to be equal to one ($||\psi|| = 1$).
%
%
Normalizing the state vector ensures that all possible measurement probabilities will always sum to exactly 100\%.
%
%

The \textbf{outer product} of $\ket{\psi}$ and $\ket{\phi}$ forms a linear operator (a matrix) denoted as $A = \ket{\psi}\bra{\phi}$.
%
%
Applying this operator to a new vector $\ket{\xi}$ yields a vector proportional to $\ket{\psi}$:
\begin{equation}
    (\ket{\psi}\bra{\phi})\ket{\xi} = \ket{\psi}(\braket{\phi}{\xi}) = c \ket{\psi}
\end{equation}
where $c = \braket{\phi}{\xi}$ is scalar multiplier.
%
%

Outer products are utilized in quantum algorithms to construct projection operators, which map directly to quantum measurements.
%
%
\subsection{Linear Operators}

Quantum states do not exist statically, as they constantly evolve.
This evolution is described by linear operators acting upon the Hilbert space $\mathcal{H}$.
%
%

An operator $A$ is considered linear if it preserves the principle of superposition, meaning 
%
%
$A(c_1\ket{\psi_1} + c_2\ket{\psi_2}) = c_1 A\ket{\psi_1} + c_2 A\ket{\psi_2}$.
%
%

\subsubsection{Hermitian Operators and Observables}

An operator $A$ is \textit{Hermitian} (or self-adjoint) if it is equal to its conjugate transpose, $A = A^\dagger$.
%
%
The Spectral Theorem states that for any Hermitian operator $M$ on a finite dimensional space, 
there inherently exists an orthonormal basis of eigenvectors $\{\ket{m_i}\}$ paired with 
purely real eigenvalues $m_i$ such that:
%
%
\begin{equation}
    M = \sum_i m_i \ket{m_i}\bra{m_i}
\end{equation}
%
%

This decomposition has a physical consequences. The result of measuring the observable $M$ will always be one of those real values $m_i$.
%
%
\subsubsection{Unitary Operators and Evolution}

An operator $U$ is \textit{unitary} if its inverse is exactly equal to its adjoint:
%
%
\begin{equation}
    U^\dagger U = U U^\dagger = I
\end{equation}
%
%

Unitary operators preserve the inner product: $\bra{\psi} U^\dagger U \ket{\phi} = \braket{\psi}{\phi}$, 
which means they preserve the norm (length) of any state vector they act upon.
%
%
This preservation is essential for quantum computing, because the norm corresponds directly to total probability.
%
%
Application of an unitary operator acts as a guarantee that the sum of probabilities for all possible outcomes remains exactly $1$ throughout the entire duration of the computation.
%
%
Therefore, all logic gates applied to a quantum register must be formulated as unitary matrices \cite{nielsen2010quantum}.